### 2023-08-24 documents/config/own_abbreviations.tex
fixes #26

\DeclareAcronym{p}{short=p,long=patient-like~argument~of~canonical~transitive~verb,short-format=\scshape}

@@

\DeclareAcronym{s}{short=s,long=single~argument~of~canonical~intransitive~verb,short-format=\scshape}

>>>>

\DeclareAcronym{p}{short=P,long=patient-like~argument~of~canonical~transitive~verb,short-format=\scshape}

@@

\DeclareAcronym{s}{short=S,long=single~argument~of~canonical~intransitive~verb,short-format=\scshape}

### 2023-08-24 documents/sections/proto_carib.tex
fixes #26

\ex
    \begin{threeparttable}
    	\begin{tabular}[t]{ll}
    		\PPar & \gl{1}\gl{sa} \rc{w-} to \gl{1}>\gl{3}\\
    		& \gl{1+2} \rc{k-} to \gl{1}\gl{sp}\tnote{a}\\
    		& \gl{1+2} \rc{kɨt-} to \gl{1+2}\gl{sp}\tnote{a}\\
    		\PWai & \gl{1}\gl{sp} \rc{k-} to \gl{1}\gl{sa}\\
    		& \rc{owɨ(ro) j-} \qu{\gl{1}\gl{pro} \gl{lk}} for \gl{1}\gl{p}\\
    		\waiwai & \gl{2}\gl{sa} \obj{m-} to \gl{2}\gl{sp}
    	\end{tabular}
    \begin{tablenotes}
    	\footnotesize
    	\item[a] Completed in \PWai, ongoing in \kaxui.
    \end{tablenotes}
    	\end{threeparttable}
\xe
%
%\begin{enumerate}
%	\item \PPar \begin{enumerate}
%		\item \gl{1}\gl{sa} \rc{w-} to \gl{1}>\gl{3}
%		\item \gl{1+2} \rc{k-} to \gl{1}\gl{sp} (completed in \PWai, ongoing in \kaxui)
%		\item \gl{1+2} \rc{kɨt-} to \gl{1+2}\gl{sp} (completed in \PWai, ongoing in \kaxui)
%	\end{enumerate}
%	\item \PWai \begin{enumerate}
%		\item \gl{1}\gl{sp} \rc{k-} to \gl{1}\gl{sa}
%		\item innovative \rc{owɨro j-} \qu{\gl{1}\gl{pro} \gl{lk}} for \gl{1}\gl{p}
%	\end{enumerate}
%	\item \waiwai \begin{enumerate}
%		\item \gl{2}\gl{sa} \obj{m-} to \gl{2}\gl{sp}\end{enumerate}
%\end{enumerate}
%
All innovations are person marker extensions except the innovative first person P form, which combined a pronoun with the linker \rc{j-}.
\cref{fig:par_ext} shows them in bold and reproduces \posscite{gildea1998} tables as a tree diagram, with adapted transcription and an additional \kaxui \gl{1}\gl{sp} marker ∅/\obj{j-} \perscommpar{Spike Gildea}.
%
\begin{figure}[hbt]
	\centering
	\setlength{\tabcolsep}{2pt}
	\fbox{\begin{tikzpicture}[
			every node/.append style={align=center},
			level distance=80pt,
			]
			\Tree[.{\PC{}\pcone}
			[.{\PPar{}\ppone}
			{\kaxui{}\kaxone}
			[.{\PWai{}\pwone}
			{\hixka{}\hixone}
			{\waiwai{}\waione}
			]
			]
			]
	\end{tikzpicture}}
	\caption[Person marker extensions in Parukotoan]{Person marker extensions in Parukotoan, after \textcite[94]{gildea1998}}
	\label{fig:par_ext}
\end{figure}
%
\hixka has preserved split-\gl{s} only in the second person prefixes, while \kaxui still shows the variation reconstructible to \PPar in its first person and \gl{1+2} prefixes.
\waiwai has lost the system entirely, which notably happened via three diachronically distinct innovations.
%In this case, the loss of the inflectional classes also entailed the loss of the other morphological traces of the split-\gl{s} system.
%That is, the \gl{sa} class marker \rc{w-} \pcref{sec:split} was lost in \waiwai, as evidenced by the contrast between the \waiwai and \kaxui deverbal forms in \exref{wailoss.wai-158} and \exref{wailoss.kax-138}.
%Similarly, the \gl{2}\gl{sp} prefix \obj{a-} was extended to imperatives of (former) \gl{sa} verbs \exref[wailoss.wai-112]{wailoss.hix-43}, but only C-initial ones \parencite[62]{waiwaihawkins1998}.

@@

\subsection{The Cariban split-\gl{s} system}
\label{sec:split}
As seen in \cref{sec:pc_person}, the split between \gl{sa} and \gl{sp} verbs was instantiated by inflection classes within the \PC \setone person paradigm, but this was not the only difference:
In deverbalized forms, \gl{sa} verbs took a class marker \rc{w-}, while \gl{sp} verbs lacked that prefix \parencites[89, 141--142]{gildea1998}[208]{meira2000split}.
Also, \gl{sp} verbs took the \gl{2}\gl{sp} prefix \rc{ə(j)-} in imperatives, while \gl{sa} verbs were unprefixed \parencites[89]{gildea1998}[208]{meira2000split}.

@@

In modern instantiations of Cariban split-\gl{s}, mismatches between the semantics of verbs and their \gl{sa} or \gl{sp} status are common, exemplified with \kalina data in \exref{kar2}.
%
%\gl{a}- or \gl{p}-oriented inflectional morphology are common, exemplified with \kalina data in \exref[kar1]{kar2}.
%
%\pex<kar1> \kalina
%\a<kar-60>
%\begingl
%\gla mi-kupi-ja//
%\glb \gl{2}>\gl{3}-bathe-\gl{prs}//
%\glft \qu{You bathe him/her.} \parencite[][160]{hoff1968carib}//
%\endgl
%\a<kar-62>
%\begingl
%\gla a-kupi-ja//
%\glb \gl{3}>\gl{2}-bathe-\gl{prs}//
%\glft \qu{S/he bathes you.} \parencite[][63]{yamada2011evidentiality}//
%\endgl
%\xe
%%
%In \exref{kar1}, the choice between the second person \gl{a}- and \gl{p}-oriented markers \obj{mi-} and \obj{a-} depends on the scenario:
%The transitive verb \obj{kupi} \qu{to bathe} takes \obj{mi-} in \gl{2}>\gl{3} scenarios \exref{kar1.kar-60}, but \obj{a-} in \gl{3}>\gl{2} scenarios \exref{kar1.kar-62}.
%The intransitive verbs in \exref{kar2} show the same person markers, but the choice of marker depends on the verb rather than semantics.
%
\input{examples/kar2.tex}
%
The \gl{sa} verb \obj{ekema} \qu{to be afraid} takes an \gl{a}-oriented marker \exref{kar2.kar-59}, while the \gl{sp} verb \obj{awomɨ} \qu{to get up} takes a \gl{p}-oriented marker \exref{kar2.kar-61}.
In both cases, the prefix does not appear to contribute to the semantics of the predicate, since there are clear mismatches:
\qu{to be afraid} with an \dbqu{agentive} marker can hardly be considered a volitional act, while  \qu{to get up} with a \dbqu{patientive} marker is clearly volitional.
\textcite{meira2000split} investigates a corpus of intransitive verbs from \trio, \kalina, \apalai, and \wayana, and categorizes them by applying different criteria commonly encountered in split-\gl{s} systems.
He shows that neither (non\-)activities, % \parencite[210]{meira2000split}, 
(non-)agency, % \citeyearpar[212]{meira2000split}, 
(in-)animacy, % \citeyearpar[213]{meira2000split},
nor Aktionsart % \citeyearpar[215]{meira2000split} 
satisfactorily predict the class membership of intransitive verbs in any of the languages.

Rather, the reason for a verb to take \gl{a}- or \gl{p}-oriented prefixes is (at least diachronically) a morphological one.
\textcite[217--221]{meira2000split} demonstrates that those intransitive verbs which (etymologically) have a derivational detransitivizing prefix are treated as \gl{sa} verbs, while essentially all others are \gl{sp} verbs:
\begin{quotebox}{\parencite[201]{meira2000split}}
	Almost all verbs in the \gl{sa} class are detransitivized forms of transitive verbs, either synchronically (with still existing transitive sources) or diachronically (with reconstructible but no longer existing transitive sources)
\end{quotebox}
%\textcite[221--223]{meira2000split} also argues that the detransitivizing prefixes are indeed deriving \gl{sa} verbs, rather than being inflectional in nature:
%\begin{inlinelist}
%	\item there are a few underived \gl{sa} verbs, with no detransitivizing prefix
%	\item \gl{sa} verbs can develop irregular semantics compared to their transitive counterparts
%	\item it is unpredictable whether the \gl{a} or \gl{p} argument of the underlying transitive verb becomes the \gl{s} of the derived \gl{sa} verb
%	\item some originally derived \gl{sa} verbs have lost their transitive counterparts
%	\item \dbqu{basic} concepts are expressed as derivations of more complex concepts, like \qu{to dance (\gl{sa})} from \qu{to dance with (\gl{tr})}
%\end{inlinelist}.
He notes that this leads to an inflectional split not based in meaning, but rather morphology:

>>>>

\ex
    \begin{threeparttable}
    	\begin{tabular}[t]{ll}
    		\PPar & \gl{1}\gl{sa} \rc{w-} to \gl{1}>\gl{3}\\
    		& \gl{1+2} \rc{k-} to \gl{1}\gl{sp}\tnote{a}\\
    		& \gl{1+2} \rc{kɨt-} to \gl{1+2}\gl{sp}\tnote{a}\\
    		\PWai & \gl{1}\gl{sp} \rc{k-} to \gl{1}\gl{sa}\\
    		& \rc{owɨ(ro) (j-)} \qu{\gl{1}\gl{pro} (\gl{lk})} for \gl{1}\gl{p}\\
    		\waiwai & \gl{2}\gl{sa} \obj{m-} to \gl{2}\gl{sp}
    	\end{tabular}
    \begin{tablenotes}
    	\footnotesize
    	\item[a] Completed in \PWai, ongoing in \kaxui.
    \end{tablenotes}
    	\end{threeparttable}
\xe
%
%\begin{enumerate}
%	\item \PPar \begin{enumerate}
%		\item \gl{1}\gl{sa} \rc{w-} to \gl{1}>\gl{3}
%		\item \gl{1+2} \rc{k-} to \gl{1}\gl{sp} (completed in \PWai, ongoing in \kaxui)
%		\item \gl{1+2} \rc{kɨt-} to \gl{1+2}\gl{sp} (completed in \PWai, ongoing in \kaxui)
%	\end{enumerate}
%	\item \PWai \begin{enumerate}
%		\item \gl{1}\gl{sp} \rc{k-} to \gl{1}\gl{sa}
%		\item innovative \rc{owɨro j-} \qu{\gl{1}\gl{pro} \gl{lk}} for \gl{1}\gl{p}
%	\end{enumerate}
%	\item \waiwai \begin{enumerate}
%		\item \gl{2}\gl{sa} \obj{m-} to \gl{2}\gl{sp}\end{enumerate}
%\end{enumerate}
%
All innovations are person marker extensions except the innovative first person P form, which consists of a pronoun, combined with the linker \rc{j-} for V-initial stems.
\cref{fig:par_ext} shows them in bold and reproduces \posscite{gildea1998} tables as a tree diagram, with adapted transcription and an additional \kaxui \gl{1}\gl{sp} marker ∅/\obj{j-} \perscommpar{Spike Gildea}.
%
\begin{figure}[hbt]
	\centering
	\setlength{\tabcolsep}{2pt}
	\fbox{\begin{tikzpicture}[
			every node/.append style={align=center},
			level distance=80pt,
			]
			\Tree[.{\PC{}\pcone}
			[.{\PPar{}\ppone}
			{\kaxui{}\kaxone}
			[.{\PWai{}\pwone}
			{\hixka{}\hixone}
			{\waiwai{}\waione}
			]
			]
			]
	\end{tikzpicture}}
	\caption[Person marker extensions in Parukotoan]{Person marker extensions in Parukotoan, after \textcite[94]{gildea1998}}
	\label{fig:par_ext}
\end{figure}
%
\hixka has preserved split-\gl{s} only in the second person prefixes, while \kaxui still shows the variation reconstructible to \PPar in its first person and \gl{1+2} prefixes.
\waiwai has lost the system entirely, which notably happened via three diachronically distinct innovations.
%In this case, the loss of the inflectional classes also entailed the loss of the other morphological traces of the split-\gl{s} system.
%That is, the \gl{sa} class marker \rc{w-} \pcref{sec:split} was lost in \waiwai, as evidenced by the contrast between the \waiwai and \kaxui deverbal forms in \exref{wailoss.wai-158} and \exref{wailoss.kax-138}.
%Similarly, the \gl{2}\gl{sp} prefix \obj{a-} was extended to imperatives of (former) \gl{sa} verbs \exref[wailoss.wai-112]{wailoss.hix-43}, but only C-initial ones \parencite[62]{waiwaihawkins1998}.

@@

\subsection{The Cariban split-\gl{s} system}
\label{sec:split}
As seen in \cref{sec:pc_person}, the split between \gl{sa} and \gl{sp} verbs was instantiated by inflection classes within the \PC \setone person paradigm, but this was not the only difference:
In nonfinite forms (nominalizations and adverbializations), \gl{sa} verbs took a prefix \rc{w-}, while \gl{sp} verbs lacked that prefix \parencites[89, 141--142]{gildea1998}[208]{meira2000split}.
Also, \gl{sp} verbs took the \gl{2}\gl{sp} prefix \rc{ə(j)-} in imperatives, while \gl{sa} verbs were unprefixed \parencites[89]{gildea1998}[208]{meira2000split}.

@@

%In modern instantiations of Cariban split-\gl{s}, mismatches between the semantics of verbs and their \gl{sa} or \gl{sp} status are common, exemplified with \kalina data in \exref{kar2}.
%
%\input{examples/kar2.tex}
%
%The \gl{sa} verb \obj{ekema} \qu{to be afraid} takes an \gl{a}-oriented marker \exref{kar2.kar-59}, while the \gl{sp} verb \obj{awomɨ} \qu{to get up} takes a \gl{p}-oriented marker \exref{kar2.kar-61}.
%In both cases, the prefix does not appear to contribute to the semantics of the predicate, since there are clear mismatches:
%\qu{to be afraid} with an \dbqu{agentive} marker can hardly be considered a volitional act, while  \qu{to get up} with a \dbqu{patientive} marker is clearly volitional.
%\gl{a}- or \gl{p}-oriented inflectional morphology are common, exemplified with \kalina data in \exref[kar1]{kar2}.
%
%\pex<kar1> \kalina
%\a<kar-60>
%\begingl
%\gla mi-kupi-ja//
%\glb \gl{2}>\gl{3}-bathe-\gl{prs}//
%\glft \qu{You bathe him/her.} \parencite[][160]{hoff1968carib}//
%\endgl
%\a<kar-62>
%\begingl
%\gla a-kupi-ja//
%\glb \gl{3}>\gl{2}-bathe-\gl{prs}//
%\glft \qu{S/he bathes you.} \parencite[][63]{yamada2011evidentiality}//
%\endgl
%\xe
%%
%In \exref{kar1}, the choice between the second person \gl{a}- and \gl{p}-oriented markers \obj{mi-} and \obj{a-} depends on the scenario:
%The transitive verb \obj{kupi} \qu{to bathe} takes \obj{mi-} in \gl{2}>\gl{3} scenarios \exref{kar1.kar-60}, but \obj{a-} in \gl{3}>\gl{2} scenarios \exref{kar1.kar-62}.
%The intransitive verbs in \exref{kar2} show the same person markers, but the choice of marker depends on the verb rather than semantics.
%

\textcite{meira2000split} investigates a corpus of intransitive verbs from \trio, \kalina, \apalai, and \wayana, and categorizes them by applying different criteria commonly encountered in split-\gl{s} systems.
He shows that neither (non\-)activities, % \parencite[210]{meira2000split}, 
(non-)agency, % \citeyearpar[212]{meira2000split}, 
(in-)animacy, % \citeyearpar[213]{meira2000split},
nor Aktionsart % \citeyearpar[215]{meira2000split} 
satisfactorily predict the class membership of intransitive verbs in any of the languages.
Rather, the reason for a verb to take \gl{a}- or \gl{p}-oriented prefixes is (at least diachronically) a morphological one.
\textcite[217--221]{meira2000split} demonstrates that those intransitive verbs which (etymologically) have a derivational detransitivizing prefix are treated as \gl{sa} verbs, while almost all others are \gl{sp} verbs:
\begin{quotebox}{\parencite[201]{meira2000split}}
	Almost all verbs in the \gl{sa} class are detransitivized forms of transitive verbs, either synchronically (with still existing transitive sources) or diachronically (with reconstructible but no longer existing transitive sources)
\end{quotebox}
%\textcite[221--223]{meira2000split} also argues that the detransitivizing prefixes are indeed deriving \gl{sa} verbs, rather than being inflectional in nature:
%\begin{inlinelist}
%	\item there are a few underived \gl{sa} verbs, with no detransitivizing prefix
%	\item \gl{sa} verbs can develop irregular semantics compared to their transitive counterparts
%	\item it is unpredictable whether the \gl{a} or \gl{p} argument of the underlying transitive verb becomes the \gl{s} of the derived \gl{sa} verb
%	\item some originally derived \gl{sa} verbs have lost their transitive counterparts
%	\item \dbqu{basic} concepts are expressed as derivations of more complex concepts, like \qu{to dance (\gl{sa})} from \qu{to dance with (\gl{tr})}
%\end{inlinelist}.
He notes that this leads to an inflectional split not based in meaning, but rather morphology:
